Pulsar Acceleration Shifts from nearby Supernova Explosion

### 2016-01-15 draft3_TC.tex
ready for arxiv
final changes ready for arctic

\documentclass[aps,showpacs,twocolumn,floats,prd,superscriptaddress,nofootinbib]{revtex4-1} 
\usepackage{graphicx,amsmath,amssymb,amstext}
\usepackage{amssymb,amsbsy,amsfonts,amsthm,color}

@@

\author{Ue-Li Pen}
\email{pen@cita.utoronto.ca}
\affiliation{Canadian Institute of Theoretical Astrophysics, 60 St George St, Toronto, ON M5S 3H8, Canada.}
\affiliation{Canadian Institute for Advanced Research, CIFAR program in Gravitation and Cosmology.}

@@

\begin{abstract}
We show that when a supernova explodes, a nearby pulsar goes through two changes in the time-derivative of its period. A stable, millisecond pulsar can allow us to measure such effect. This may improve our measurement on the total energy released in neutrinos and also the actual orientation of the supernova-pulsar system. A joint timing of a few pulsars can even triangulate the location of a hidden supernova.
\end{abstract}

@@

Since this neutrino shell is very thin---passing through either the Earth or the pulsar in a time much shorter than the typical duration for timing observations---we can treat it as a co-dimension-1 delta-function and describe the spacetime by Israel Junction Conditions (IJC) \cite{Isr66}. Furthermore, we will treat both the neutrinos and the pulsar signal as propagating at the speed of light. In reality that is not exactly true, with photons moving even slower than neutrinos. Nevertheless, the time delay is again much shorter compared to the duration of timing observations, thus it does not invalidate our result. The analytical model then becomes extremely simple, since it involves nothing but tracking geodesics\cite{BouFre07,JohYan10}.

@@

Our conclusion above is exactly correct in full nonlinear general relativity, although it might be surprising to some readers. Here we will expand the two steps separately to leading order for better physical intuitions. First of all, it is quite realistic to set $r_e \gg r_p$ and treat it as being infinite, so the only physical change happens in the pulsar.

Instead of using Eq.~(\ref{eq-JuncPuls}), if we had assumed that the pulsar is at rest in both geometries before and after the neutrino shell crosses, then we would have derived a frequency ratio of $[1-\Delta M/r_p]$ between these two photons\footnote{In fact, any non-relativistic initial velocity leads to this result.}. This would have reflected the fact that they climbed out of two different gravitational potentials, but it cannot be the full story. If the pulsar was at rest before the neutrino shell hits it, Eq.~(\ref{eq-JuncPuls}) demands that it picks up a velocity in the new coordinate
 \begin{eqnarray}
p_\mu &\equiv& \left( \left( 1 - \frac{2M_i}{r} \right)^{-\frac{1}{2}}, 0, 0,0\right)_{\rm i} 
\label{eq-PulsarRest} \\ \nonumber  &\equiv&	
\left( \left( 1 - \frac{2M_f}{r} \right)^{-\frac{1}{2}}, -\frac{\Delta M}{r_p},0,0 \right)_{\rm f}~.
\end{eqnarray}
The new velocity, $v=\Delta M/r_p$, is falling toward the SN remnant. This then creates a Doppler shift that exactly cancels the difference in gravitational potential, thus netting no frequency change.\footnote{This change in coordinate velocity is not due to any direct interaction with the neutrino shell, but simply an artifact of coordinate change across the junction.}
\begin{equation}
(w_e)_{\rm i} \equiv w_p \left(1-\frac{M_i}{r_p}\right) = 
w_p (1-v)\left(1-\frac{M_f}{r_p}\right) \equiv (w_e)_{\rm f}~.
\label{eq-leading}
\end{equation}

@@

Let us continue the approximation in the previous section and take a time derivative on Eq.~(\ref{eq-leading}). Realistically, $v \ll1$ and $r_p\gg M$, then the dominant change in time comes from the velocity.
\begin{equation}
\frac{\dot{w_e}}{w_e} = -\dot{v} = -\frac{M}{r_p^2}~.
\label{eq-AccChange}
\end{equation}
When the pulsar crosses the neutrino shell, the enclosed mass certainly changes from $M_i$ to $M_f$, and $\dot{w_e}$ indeed changes suddenly. It is not a suddenly observable change, since the change in $w_e$ needs some time to build up before it is observable. Nevertheless, this is the leading order observable for us.

@@

Right before $t_1$, we can find a photon that travels from the pulsar to the earth totally in the ``initial" geometry. Without loss of generality, we let the pulsar be at rest. Its 4-vector is then given by the first expression in Eq.~(\ref{eq-PulsarRest}). As shown in Fig.\ref{fig:3}, a photon aiming toward the earth must have $b = r_p\sin\theta$. We can then solve for $(w_e)_{t_1} \approx (w_\infty)_{t_1}$ by
\begin{eqnarray}
w_p = -(p_\mu k^\mu)_{\rm i} &=& 
(w_\infty)_{t_1} \left(1-\frac{2M_i}{r_p}\right)^{-1/2}~, \nonumber \\
(w_e)_{t_1} &\approx& w_p \left(1-\frac{M_i}{r_p}\right)~.
\end{eqnarray}
Right after $t_2$, a photon goes from the pulsar to the earth entirely in the ``final" geometry, and $p_\mu$ is given by the second expression of Eq.~(\ref{eq-PulsarRest}). We can similarly solve for $(w_e)_{t_1}\approx (w_\infty)_{t_2}$. 
\begin{eqnarray}
%w_p = -(p_\mu k^\mu)_f 
%&=& (w_\infty)_{t_2} \left[ \left(1-\frac{2M_f}{r_p}\right)^{-1/2} + \sqrt{1-\frac{b^2}{r_p^2 }\left(1-\frac{2M_f}{r_p}\right)} v \left(1-\frac{2M_f}{r_p}\right)^{-1}\right] 
%\nonumber \\
(w_e)_{t_2} &\approx& w_p \left(1-\frac{M_f}{r_p} - \cos\theta \frac{\Delta M}{r_p}\right)~.
\end{eqnarray}
Note that this $p_\mu$ only includes the change due to crossing the shell, without the result of actual acceleration from the SN's gravity. Since the later is already taken into account in the chart of Sec.\ref{sec-3d}, we are indeed calculating the extra contribution to
\begin{equation}
\frac{\dot{w_e}}{w_e}\bigg|_{\rm extra \ from \ t_1 \ to \ t_2} = \frac{(w_e)_{t_2}-(w_e)_{t_1}}{\Delta t} = \frac{\Delta M}{r_p^2}~.
\end{equation}
Interestingly, there is no $\theta$ dependence here.

@@

That means, if we observe three different values of $(\dot{w_e}/w_e)$ in the three stages, we can subtract the value in the first stage and get two independent data corresponding to the effect we calculated.
\begin{center}
\begin{tabular}{| c | c | c |} 
\hline
                & $t_1<t<t_2$ & $t_2<t$ \\ 
       \hline 
$\Delta(\dot{w_e}/w_e)$ & $ \Delta M/r_p^2$ 
& $(\Delta M/r_p^2)\cos\theta$   \\ 
\hline 
\end{tabular}
\end{center}

@@

In practice $\dot{w_e}$ is measured by allowing the signal to build up a frequency change, and then to build up again in the phase change. That means within the duration of the middle stage, the arrival time of the pulse changes by 
\begin{equation}
\delta T_{\rm arrival} \sim \Delta\left(\frac{\dot{w_e}}{w_e}\right)(\Delta t)^2
\sim \Delta M~.
\end{equation}
Interestingly, the $1/r_p^2$ in gravitational force exactly cancels with the accumulation time $(\Delta t)^2$. That means independent of $r_p$, a pulsar can detect this change as long as the timing accuracy is better than the light-crossing time of the effective Schwarzschild radius of the neutrino shell\footnote{In practice, we still want the pulsar to be somewhere close to the SN. Without that condition, we cannot be certain that there are no other comparable changes to acceleration during $\Delta t$.}. We can estimate $\Delta M\sim$ a fraction of solar mass $\sim10^{-6}$ seconds. $\delta T_{\rm arrival}$ for a stable, millisecond pulsar is already better \cite{PulsarTiming}\footnote{Our conservative estimation assumes only two measurements: in the beginning and the end of the duration $\Delta t$. In practice, one has $N$ measurements during the process, which increases the accuracy by another factor of $\sqrt{N}$.}.

With the forecast of thousands of millisecond pulsars to be found by the Square Kilometer Array \cite{SKA}, a new SN effectively sits within a timing-grid. At the instant the SN is detected at earth, even if only from direct neutrino detection, we know that all pulsar timing periodic derivatives change, and by different amounts depending on their distances. The closest or the best timed pulsars may detect such effect within a decade or two, and others gradually follow. Such a joint timing can triangulate the location of the SN and prepare us to detect the second acceleration change. Given a few SN per century\cite{SNrate06} an actual detection is possible in the foreseeable future.

@@

We thank xxx useful discussions. We are supported in part by xxx fundings.

>>>>

\documentclass[aps,showpacs,twocolumn,floats,prd,superscriptaddress,nofootinbib]{revtex4} 
\usepackage{graphicx,amsmath,amssymb,amstext}
\usepackage{amssymb,amsbsy,amsfonts,amsthm,color}

@@

\author{Ue-Li Pen}
\email{pen@cita.utoronto.ca}
\affiliation{Canadian Institute of Theoretical Astrophysics, 60 St George St, Toronto, ON M5S 3H8, Canada.}
\affiliation{Canadian Institute for Advanced Research, CIFAR program in Gravitation and Cosmology.}
\affiliation{Perimeter Institute of Theoretical Physics, 31 Caroline Street North, Waterloo, ON N2L 2Y5, Canada.}

@@

\begin{abstract}
We show that when a supernova explodes, a nearby pulsar goes through two changes in the time-derivative of its observed period. A stable, millisecond pulsar can allow us to measure such effect. This may improve our measurement on the total energy released in neutrinos and also the orientation of the supernova-pulsar system. 
\end{abstract}

@@

Since this neutrino shell is very thin---passing through either the Earth or the pulsar in a time much shorter than the typical duration for timing observations---we can treat it as a co-dimension-1 delta-function and describe the spacetime by Israel Junction Conditions (IJC) \cite{Isr66}. Furthermore, we will treat both the neutrinos and the pulsar signal as propagating at the speed of light. In reality that is not exactly true, with photons moving even slower than neutrinos. Nevertheless, the time delay is again much shorter compared to the duration of timing observations, thus it does not invalidate our result. The analytical model then becomes quite simple, since it involves just tracking geodesics across thin shells \cite{BouFre07,JohYan10}.

@@

Our conclusion above is exactly correct in full nonlinear general relativity, although it might be surprising to some readers. Here we will expand the two steps separately to leading order for better physical intuitions. First of all, it is quite realistic to set $r_e \gg r_p$ and treat it as being infinite, so the only physical change happens at the pulsar.

Instead of using Eq.~(\ref{eq-JuncPuls}), if we had assumed that the pulsar is at rest in both coordinates before and after the neutrino shell crosses, then we would have derived a frequency ratio of $[1-\Delta M/r_p]$ between these two photons\footnote{In fact, any non-relativistic initial velocity leads to this result.}. This would have reflected the fact that they climbed out of two different gravitational potentials, but it cannot be the full story. If the pulsar was at rest before the neutrino shell hits it, Eq.~(\ref{eq-JuncPuls}) demands that it picks up a velocity in the new coordinate
 \begin{eqnarray}
p_\mu &\equiv& \left( \left( 1 - \frac{2M_i}{r} \right)^{-\frac{1}{2}}, 0, 0,0\right)_{\rm i} 
\label{eq-PulsarRest} \\ \nonumber  &\equiv&	
\left( \left( 1 - \frac{2M_f}{r} \right)^{-\frac{1}{2}}, -\frac{\Delta M}{r_p},0,0 \right)_{\rm f}~.
\end{eqnarray}
The new velocity, $v=\Delta M/r_p$, is falling toward the SN remnant. This then creates a Doppler shift that exactly cancels the difference in gravitational potential, thus netting no frequency change.\footnote{This change in coordinate velocity is not due to any direct interaction with the neutrino shell, but simply an artifact of coordinate change across the junction.}
\begin{equation}
(w_e)_{\rm i} \equiv w_p \left(1-\frac{M_i}{r_p}\right) = 
w_p (1-v)\left(1-\frac{M_f}{r_p}\right) \equiv (w_e)_{\rm f}~.
\label{eq-leading}
\end{equation}

@@

Let us continue the approximation in the previous section and take a time derivative on Eq.~(\ref{eq-leading}). Realistically, $v \ll1$ and $r_p\gg M$, then the dominant change in time comes from the velocity.
\begin{equation}
\frac{\dot{w_e}}{w_e} = -\dot{v} = -\frac{M}{r_p^2}~.
\label{eq-AccChange}
\end{equation}
When the pulsar crosses the neutrino shell, the enclosed mass certainly changes from $M_i$ to $M_f$, and $\dot{w_e}$ indeed changes suddenly. It is not a suddenly observable change, since it needs some time to build up the change in $w_e$ and then in the phase before it is observable. Nevertheless, this is the leading order observable for us.

@@

Right before $t_1$, we can find a photon that travels from the pulsar to the earth totally in the ``initial" geometry. Without loss of generality, we let the pulsar be at rest. Its 4-vector is then given by the first expression in Eq.~(\ref{eq-PulsarRest}). As shown in Fig.\ref{fig:3}, a photon aiming toward the earth must have $b = r_p\sin\theta$. We can then solve for $(w_e)_{t_1} \approx (w_\infty)_{t_1}$ by
\begin{eqnarray}
w_p = -(p_\mu k^\mu)_{\rm i} &=& 
(w_\infty)_{t_1} \left(1-\frac{2M_i}{r_p}\right)^{-1/2}~, \nonumber \\
(w_e)_{t_1} &\approx& w_p \left(1-\frac{M_i}{r_p}\right)~.
\end{eqnarray}
Right after $t_2$, a photon goes from the pulsar to the earth entirely in the ``final" geometry, and $p_\mu$ is given by the second expression of Eq.~(\ref{eq-PulsarRest}). We can similarly solve for $(w_e)_{t_1}\approx (w_\infty)_{t_2}$. 
\begin{eqnarray}
%w_p = -(p_\mu k^\mu)_f 
%&=& (w_\infty)_{t_2} \left[ \left(1-\frac{2M_f}{r_p}\right)^{-1/2} + \sqrt{1-\frac{b^2}{r_p^2 }\left(1-\frac{2M_f}{r_p}\right)} v \left(1-\frac{2M_f}{r_p}\right)^{-1}\right] 
%\nonumber \\
(w_e)_{t_2} &\approx& w_p \left(1-\frac{M_f}{r_p} - \cos\theta \frac{\Delta M}{r_p}\right)~.
\end{eqnarray}
Note that this $p_\mu$ only includes the change due to crossing the shell, without the result of actual acceleration from the SN's gravity. Since the later is already taken into account in the chart of Sec.\ref{sec-3d}, we are indeed calculating the extra contribution to
\begin{equation}
\frac{\dot{w_e}}{w_e}\bigg|_{\rm extra \ from \ t_1 \ to \ t_2} = \frac{(w_e)_{t_2}-(w_e)_{t_1}}{w_e\Delta t} = \frac{\Delta M}{r_p^2}~.
\end{equation}
Interestingly, there is no $\theta$ dependence here.

@@

That means, if we observe three different values of $(\dot{w_e}/w_e)$ in the three stages, then the two sharp changes of acceleration should correspond to the effect we calculated.
\begin{center}
\begin{tabular}{| c | c | c |} 
\hline
                & at $t_1$ & at $t_2$ \\ 
       \hline 
$\Delta(\dot{w_e}/w_e)$ & $ \Delta M/r_p^2$ 
& $(\Delta M/r_p^2)(\cos\theta-1)$   \\ 
\hline 
\end{tabular}
\end{center}
The two changes always have opposite signs. The net change in the end is always $(\Delta M/r_p^2)\cos\theta$, which is the projected change of acceleration.

@@

In practice $\dot{w_e}$ is measured by allowing the signal to build up a frequency change, and then to build up again in the phase change. That means within the duration of the middle stage, the arrival time of the pulse changes by 
\begin{equation}
\delta T_{\rm arrival} \sim \Delta\left(\frac{\dot{w_e}}{w_e}\right)(\Delta t)^2
\sim \Delta M~.
\end{equation}
Interestingly, the $1/r_p^2$ in gravitational force exactly cancels with the accumulation time $(\Delta t)^2$. That means independent of $r_p$, a pulsar can detect this change as long as the timing accuracy is better than the light-crossing time of the effective Schwarzschild radius of the neutrino shell\footnote{In practice, we still want the pulsar to be somewhere close to the SN. Without that condition, we cannot be certain that there are no other comparable changes to acceleration during $\Delta t$.}. We can estimate $\Delta M\sim$ a fraction of solar mass $\sim10^{-6}$ seconds. The timing accuracy for a stable, millisecond pulsar is already better \cite{PulsarTiming}\footnote{Our conservative estimation assumes only two measurements: in the beginning and the end of the duration $\Delta t$. In practice, one has $N$ measurements during the process, which increases the accuracy by another factor of $\sqrt{N}$.}.

With the forecast of thousands of millisecond pulsars to be found by the Square Kilometer Array \cite{SKA}, a new SN effectively sits within a timing-grid. At the instant the SNe is detected at earth, even if only from direct neutrino detection, we know that the derivatives of all pulsars' periods change, and by different amounts depending on their distances. The closest or the best timed pulsars may detect the first change within a decade or two. The second change will also happen around the same time scale for those with small $\theta$. Given a few SNe per century \cite{SNrate06}, a signature detection of this two-consecutive-acceleration-shifts is possible in the foreseeable future.

@@

We thank John Antoniadis for inspiring discussions. This work is supported by the Canadian Government through the Canadian Institute for Advance Research and Industry Canada, and by Province of Ontario through the Ministry of Research and Innovation.
